
# Prerequisites

```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

## Learning Objectives

This chapter will cover the prerequisites for this course, including:

- Installing Docker
- Installing R Studio
- Downloading data files

## Docker

For the purpose of this course, we will be using Docker to run pVACseq and
pVACfuse.
Docker is a tool that is used to automate the deployment of applications
in lightweight containers so that applications can work efficiently in
different environments in isolation. We provide versioned Docker containers
for all pVACtools releases.

In order to use Docker, you will to download the [Docker Desktop software](https://www.docker.com/get-started/).
Please ensure you select the correct install package for your operating
system.

## Terminal

We will be running Docker from the command line on your preferred terminal
using the Docker command line interface (CLI). The Docker CLI is already
included with Docker Desktop. Most operating systems already
come with a Terminal application. If yours doesn't, you will need to first
install one.

## R Studio and R package dependencies

In order to use pVACview, you will need to download R. Please refer
[here](https://cran.rstudio.com/) for downloading R (version 3.5 and above
required). You may also take the additional step of [downloading R
studio](https://www.rstudio.com/products/rstudio/download/) if
you are not familiar with launching R Shiny from the command line.

Additionally, there are a number of packages you will need to install in your R/R studio:

```{r, eval = FALSE}
install.packages("shiny", dependencies=TRUE)
install.packages("ggplot2", dependencies=TRUE)
install.packages("DT", dependencies=TRUE)
install.packages("reshape2", dependencies=TRUE)
install.packages("jsonlite", dependencies=TRUE)
install.packages("tibble", dependencies=TRUE)
install.packages("tidyr", dependencies=TRUE)
install.packages("plyr", dependencies=TRUE)
install.packages("dplyr", dependencies=TRUE)
install.packages("shinydashboard", dependencies=TRUE)
install.packages("shinydashboardPlus", dependencies=TRUE)
install.packages("fresh", dependencies=TRUE)
install.packages("shinycssloaders", dependencies=TRUE)
install.packages("RCurl", dependencies=TRUE)
install.packages("curl", dependencies=TRUE)
install.packages("string", dependencies=TRUE)
install.packages("shinycssloaders", dependencies=TRUE)
```

## Data

For this course, we have put together a set of input data for the HCC1395
cell line. Data from this cell line is commonly used as test data in bioinformatics
applications.


```{r, engine = 'bash', eval = FALSE}
TODO: download instructions
```

This course will not cover the required pre-processing steps for the pVACtools
input data but extensive instructions on how to prepare your own data for use
with pVACtools can be found at[pvactools.org](https://www.pvactools.org)
