
# Review Your Results in pVACview

```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

## Learning Objectives

This chapter will cover:

- Introduction to the pVACview module
- How to start pVACview
- How to load your pVACseq data into pVACview
- The pVACview user interface
- How to re-tier results in pVACview
- How to add comments to the variants in the report
- How to export the reviewed results

## Introduction to the pVACview module

pVACview is a R shiny based tool designed to aid specifically in the prioritization and selection of
neoantigen candidates for personalized cancer vaccines or other applications. It takes as inputs a
pVACseq output aggregate report file (tsv format) and a corresponding pVACseq output metrics file (json).
pVACview allows the user to launch an R shiny application to load and visualize the given neoantigen
candidates with detailed information including that of the genomic variant, transcripts covering the
variant, and strong-binding peptides predicted from the respective transcripts. It also incorporates
anchor prediction data for a range of class I HLA alleles and peptides ranging from 8- to 11-mers. By
taking all these types of information into account for the neoantigen candidates, researchers will be
able to make more informed decisions when deciding final peptide candidates for experiments, personalized
cancer vaccines, or T cell therapies designed to target neoantigens.

## Starting pVACview

The pVACview R source code is distributed with every pVACseq run the in
MHC_Class_I and/or MHC_Class_II subdirectories, depending on which prediction
algorithms were run. For the HCC1395 pVACseq example run you did earlier, it
can be started by running the following command in your Terminal:

```{r, engine = 'bash', eval = FALSE}
pvacview run ${PWD}/pVACtools_outputs/MHC_Class_I
```

This starts a process for the pVACview R shiny application. Do not close the
terminal window or end this process. You can now open pVACview in your browser
by navigating to [http://127.0.0.1:3333/](http://127.0.0.1:3333/).

## Uploading pVACseq result data into pVACview

The starting screen shows an upload form. You will want to upload the Class I
aggregated.tsv report under "1. Neoantigen Candidate Aggregate Report (tsv
required)". We generally recommend using Class I as the main report, as these
are more robust than Class II. However, either can be uploaded here as long as
the correct data type is selected under "Does this aggregate report file
correspond to Class I or Class II prediction data?"

Upload the matching Class I metrics.json file under "2. Neoantigen Candidate
Metrics file (json required)". The Class II aggregated.tsv can now be uploaded
as a supplementary file under "3. Additional Neoantigen Candidate Aggregate
Report (tsv required)".

This interface also optionally allows users to upload a gene-of-interest TSV
file to, e.g. highlight variants on known cancer genes. We have provided such
a TSV with your original HCC1395 downloaded data, called
"cancer_census_hotspot_gene_list.tsv".

After uploading all files, you can visualize the result by clicking on the
"Visualize" button at the bottom of the page, or the "Visualize and Explore"
tab in the sidebar.

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "pVACview data upload interface."}
ottrpal::include_slide("https://docs.google.com/presentation/d/1uz39zaObDGKhEVCGzO0JO35CTbC0oRAM0mxgLcMAA9Y/edit#slide=id.g25ad9ce8c9b_0_2")
```

## The pVACview User Interface

### Aggregate Report of Best Candidates by Variant

The main table in the Aggregate Report of Best Candidates by Variant panel
shows the best neoantigen candidate for each variant.
It lists the gene and amino acid change of the variant as well as additional
information about the best peptide and the best transcript coding for it. These
include, from left to right, whether or not the transcript is considered a good
transcript considering the `--transcript-prioritization-strategy` and
`--maximum-transcript-support-level`, the best-binding HLA
allele, the mutated position(s) of the best peptide, any positions in the peptide
where the amino acid might be problematic for manufacturing according to the
`--problematic-amino-acids`, the total number of neoantigen candidates resulting
from this variant that are included with additional detailed metadata, and the
total number of neoantigen candidates passing the binding affinity threshold set
by the user. If a gene of interest list was uploaded, variants on those genes have
their gene highlighted with a green border.

Next, this table lists the IC50 peptide MHC binding affinity for both the Best
Peptide (MT) and the matched wild type (WT). It then lists the combined mutant
percentile score, the mutant binding percentile, the mutant presentation percentile,
and the mutant immunogenicity percentile.These values are either a median of all of
the prediction scores made by the predictors selected in your pVACseq run, or the
lowest score. This depends on the value set for the `--top-score-metric` in your
run. A heatmap coloring is applied to make it easier to visually identify
well-binding peptides.

The next few columns show the VAF, coverage, and expression of the best transcript with
a bar plot background to represent where specific values fall across the entire
patient sample.

The Ref Match column reflects whether or not the best peptide was found in the
reference proteome which is undesired since such peptides are not novel and
including them in a vaccine might lead to an auto immune response.

The Tier column represents the tier assigned to the best peptide. The neoantigen
candidates in this view were mostly sorted into the Pass tier but tiers such as Low
Expression or Subclonal are applied to easily identify why a neoantigen candidate
might be unsuitable for vaccine selection. The bottom two variants in the view are
assigned the PoorBinder tier because of the binding affinity exceeding the desired
binding threshold. This is also visually indicated by the yellow cell backgrounds
and the red border around the IC50 MT value cells for these variants.

Users are able to set a status for each candidate by clicking the appropriate
buttons: thumbs-up (accept), thumbs-down (reject), or flag (requires further review).

Clicking on the row for a variant will select it. This
will update the lower panels with details for the selected variant.

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "Upon successfully uploading the relevant data files, you can explore the different aspects of your neoantigen candidates."}
ottrpal::include_slide("https://docs.google.com/presentation/d/1uz39zaObDGKhEVCGzO0JO35CTbC0oRAM0mxgLcMAA9Y/edit#slide=id.g3e3d230b076_0_0")
```

### Variant Information

The Variant Information panel shows more variant-level details of the selected
neoantigen candidate.

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "The Variant Information tab shows more details for the selected variant."}
ottrpal::include_slide("https://docs.google.com/presentation/d/1uz39zaObDGKhEVCGzO0JO35CTbC0oRAM0mxgLcMAA9Y/edit#slide=id.g3a65970eb36_0_12")
```

On the left is a three-tab section. The first tab, "Transcript Sets of Selected
Variant", shows a list of transcript sets. pVACtools bins transcripts that code for
the same set of neoantigen candidates into a set
because the neoantigen-level information for all transcripts in a set will be
identical. The transcript set containing the Best Peptide is highlighted in
green. This table shows the number of transcripts in each set, the number of
well-binding peptides the transcripts in the set code for, and the total
expression of all the transcripts in the set.

The second tab, titled Reference Matches, shows the details for any reference matches of the neoantigen
candidate. It repeats information about the Best Peptide sequence, the amino acid
change, the mutated position, and the gene for easy reference. The mutated
positions are notated in red in the Best Peptide sequence and the Query
Sequence. The Query Sequence is a longer sequence around the somatic mutation.
The Best Peptide subsequence is highlighted in yellow in the Query Sequence. For
any 8mer subsequence of the query sequence, pVACtools looks for matches in the
reference proteome. Any matches found are reported in the Hits table

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "The Reference Matches tab shows details of reference matches of the neoantigen candidate."}
ottrpal::include_slide("https://docs.google.com/presentation/d/1uz39zaObDGKhEVCGzO0JO35CTbC0oRAM0mxgLcMAA9Y/edit#slide=id.g25ad9ce8c9b_0_20")
```

The "Additional Data" tab shows the additional data for the variant if a Additional Neoantigen Candidate Aggregate
Report was uploaded. It shows the Best Peptide and its information for this variant from the
additional report. This can be used, e.g., when a Class I neoantigen candidate
is a bad binder but all other metrics look good. Oftentimes this variant can
be rescued by considering the best Class II neoantigen candidate instead.

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "The Additional Data shows data from the Additional Aggregate Report for the variant."}
ottrpal::include_slide("https://docs.google.com/presentation/d/1uz39zaObDGKhEVCGzO0JO35CTbC0oRAM0mxgLcMAA9Y/edit#slide=id.g25ad9ce8c9b_0_26")
```

The next section, "Variant & Gene Info", shows coverage and expression information as well as
the genomic coordinates for the variant.

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "The Variant & Gene INfo section show coverage and expression information for the variant."}
ottrpal::include_slide("https://docs.google.com/presentation/d/1uz39zaObDGKhEVCGzO0JO35CTbC0oRAM0mxgLcMAA9Y/edit#slide=id.g25ad9ce8c9b_0_75")
```

The last section, "Peptide Evaluation Overview", shows counts for how many peptides have been accepted,
rejected, or marked for review. For most vaccines a certain minimum number of
neoantigen candidates is desired so this panel makes it easy to review how
many neoantigen candidates are still needed.

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "The Peptide Evaluation Overview section shows how many neoantigen candidates have been accepted, rejected, marked for review, or are pending."}
ottrpal::include_slide("https://docs.google.com/presentation/d/1uz39zaObDGKhEVCGzO0JO35CTbC0oRAM0mxgLcMAA9Y/edit#slide=id.g3a65970eb36_0_18")
```

### Transcript Set Detailed Data

When selecting a transcript set in the Variant Information panel, this panel will
show details about the neoantigen candidates the transcripts in the set code
for and as well as details on the transcripts themselves.

The Peptide Candidates from Selected Transcript Set tab shows a list of
mutant and matched wildtype peptides and their IC50 binding
affinity to the patient HLA alleles. Only neoantigen candidates where at least
one peptide-MHC binding prediction falls within the  `--aggregate-inclusion-threshold`
will be shown in this table. However, for variants resulting in a large number
of well-binding candidates (e.g. frameshift variant with a long downstream
sequence), the number of included candidates will be limited to the best
`--aggregate-inclusion-count-limit` candidates. For HLA alleles where the peptide is not
well-binding the prediction details will show `X`. This table also shows the
mutant position, whether or not the neoantigen candidate has any problematic
positions, and whether or not it failed the anchor criteria. This helps in
determining whether a neoantigen candidate was deprioritized when selecting the
Best Peptide. The Best Peptide is highlighted in green.

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "The Transcript Set Detailed Data panel shows binding prediction details for the neoantigens the transcripts in the set code for."}
ottrpal::include_slide("https://docs.google.com/presentation/d/1uz39zaObDGKhEVCGzO0JO35CTbC0oRAM0mxgLcMAA9Y/edit#slide=id.g3a65970eb36_0_24")
```

The Anchor Heatmap tab shows a heatmap overlaying each included neoantigen
candidate from the selected transcript set. A darker color represents a higher
probability that a position in the peptide is an anchor. Mutated positions are
represented by red letters. More information
about how to interpret the heatmap can be found in the graph on the right of this
panel.

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "The Anchor heatmaps show which positions in a peptide are likely to be anchors."}
ottrpal::include_slide("https://docs.google.com/presentation/d/1uz39zaObDGKhEVCGzO0JO35CTbC0oRAM0mxgLcMAA9Y/edit#slide=id.g3a65970eb36_0_30")
```

This tab also includes a table listing the anchor positions for each allele
and length combination. It also includes a table listing the
per-allele-and-length anchor weights at each position.

The Transcripts in Set tabs shows details of the transcripts in the selected
set such as the transcript Ensembl ID, the the transcript expression, the
transcript's MANE Select status, the transcript's canonical status, the
transcript support level, the biotype, whether the transcript has any CDS
Flags, and the transcript length. This reflects the criteria used in determining
the Best Transcript. The Best Transcript is highlighted in green.

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "The Transcripts in Set tab shows details about the transcripts in the set."}
ottrpal::include_slide("https://docs.google.com/presentation/d/1uz39zaObDGKhEVCGzO0JO35CTbC0oRAM0mxgLcMAA9Y/edit#slide=id.g3a65970eb36_0_36")
```

### Additional Peptide Information

In the Additional Peptide Information panel, users can see more information
for the neoantigen candidate selected in the Transcript Set Detailed Data
panel.

The first tab, %ile Plot, shows a violing plot for the predicted percentile
scores for each prediction algorithm for both the neoantigen candidate and
its matched wildtype peptide. This can be used to check concordance of
predictions between the different algorithms. Each data point is colored
depending on the type of algorithm, blue for binding algorithms, yellow for
presentation algorithms, and purple for immunogenicity algorithms. Users may
filter to plot to only show a subset of algorithms by selecting the desired
algorithm type in the "Specify what data to show" dropdown.

A solid black line is used to represent the median score. A green solid line
is used to show the 2% cutoff and dashed line for the 0.5% cutoff.

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "The %ile Plot tab shows violin plots of the percentile score predicted by each algorithm."}
ottrpal::include_slide("https://docs.google.com/presentation/d/1uz39zaObDGKhEVCGzO0JO35CTbC0oRAM0mxgLcMAA9Y/edit#slide=id.g3e422d3441b_0_0")
```

The "IC50 Plot" tab shows a similar violin plot but for the IC50 binding affinity
scores as opposed to the percentile. A solid line is also used here to represent
the median score. A green solid line is used to show the 1000nM cutoff and dashed
line for the 500nM cutoff.

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "The IC50 Plot tab shows violin plots of the binding affinity scores predicted by each binding affinity algorithm."}
ottrpal::include_slide("https://docs.google.com/presentation/d/1uz39zaObDGKhEVCGzO0JO35CTbC0oRAM0mxgLcMAA9Y/edit#slide=id.g3e422d3441b_0_5")
```

The next tab, "Binding Data", shows the IC50 binding affinity and
binding percentile scores but in table format. Heatmap table cell backgrounds
indicate where each value falls in relation to binding threshold and
binding percentile threshold, respectively. Users may switch between the two
by changing the selection in the "Cell heatmap background coloring" dropdown.

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "The Binding Data tab shows a table of the IC50 binding affinity and percentile values predicted by each algorithm."}
ottrpal::include_slide("https://docs.google.com/presentation/d/1uz39zaObDGKhEVCGzO0JO35CTbC0oRAM0mxgLcMAA9Y/edit#slide=id.g3e422d3441b_0_10")
```

The Presentation Data tab shows the predicted presentation scores and presentation
percentiles. Heatmap table cell backgrounds indicate where each percentile value
falls in relation to presentation percentile threshold.

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "The Presentation Data tab shows presentation prediction scores and precentiles for the selected peptide."}
ottrpal::include_slide("https://docs.google.com/presentation/d/1uz39zaObDGKhEVCGzO0JO35CTbC0oRAM0mxgLcMAA9Y/edit#slide=id.g3e422d3441b_0_15")
```

The Immunogenicity Data tab shows the predicted immunogenicity scores and imunogenicity
percentiles. Heatmap table cell backgrounds indicate where each percentile value
falls in relation to immunogenicity percentile threshold.

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "The Immunogenicity Data tab shows immunogenicity prediction scores and precentiles for the selected peptide."}
ottrpal::include_slide("https://docs.google.com/presentation/d/1uz39zaObDGKhEVCGzO0JO35CTbC0oRAM0mxgLcMAA9Y/edit#slide=id.g3e422d3441b_0_20")
```

## Regenerate Tiers with Custom Parameters

During review of your data it might become apparent that different tiering
thresholds would've been more appropriate. pVACview allows you to re-tier your
data with custom parameters by adjusting the sliders and inputs in the
"Advanced Options: Regenerate Tiering with different parameters" panel and
pressing the "Recalculate Tiering with new parameters" button. The tiering
thresholds currently applied to your data are listed in the "Current Parameters
for Tiering" panel.

The parameters that were used in the original pVACseq run can still be viewed
in the "Original Parameters for Tiering" panel and the tiers can be reset to
those parameters by pressing the "Reset to original parameters" button.

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "Users can re-tier the neoantigen candidates by adjusting the tiering thresholds."}
ottrpal::include_slide("https://docs.google.com/presentation/d/1uz39zaObDGKhEVCGzO0JO35CTbC0oRAM0mxgLcMAA9Y/edit#slide=id.g3e422d3441b_0_31")
```

## Adding Comments to Variants

When reviewing neoantigen candidates in pVACseq, users are able to add
comments on each variant, for example, describing what additional review is
necessary. Please note that comments are not saved until the "Update Comment
Section" button is pressed.


```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "Users can leave comments on each variant."}
ottrpal::include_slide("https://docs.google.com/presentation/d/1uz39zaObDGKhEVCGzO0JO35CTbC0oRAM0mxgLcMAA9Y/edit#slide=id.g25ad9ce8c9b_0_93")
```

## Exporting the Aggregated Table

Once review of the neoantigen candidates has been completed, the results can
be exported by switching to the Export interface via the sidebar. This will
export the Aggregated Report with the updated Evaluation column and comments
added. The report can be exported in either TSV or Excel format.

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "Users can export the neoantigen candidate table after review has been completed."}
ottrpal::include_slide("https://docs.google.com/presentation/d/1uz39zaObDGKhEVCGzO0JO35CTbC0oRAM0mxgLcMAA9Y/edit#slide=id.g3e422d3441b_0_38")
```
